---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Set the working directory, read the zipped file, convert the dates to the proper format:
```{r read_file,  echo=TRUE, eval=TRUE}
options("scipen"=8)
setwd("~/Sync/Coursera/ReproResearch/RepData_PeerAssessment1")
activity <- read.csv(unzip("activity.zip", file="activity.csv"), header=TRUE, sep=",")
str(activity)
activity$date <- as.Date(activity$date)
summary(activity)
```

## What is mean total number of steps taken per day?
1. Make a histogram of the steps per day, and find the mean and median:
```{r steps_per_day,  echo=TRUE, eval=TRUE}
library(dplyr)
stepsPerDay <- activity %>%
        group_by(date) %>%
        summarise(steps_per_day = sum(steps, na.rm=TRUE))

library(ggplot2)
ggplot(data=stepsPerDay, aes(x=steps_per_day)) +
        geom_histogram() +
        ggtitle("Steps per day in the activity data")

steps.mean <- round(mean(stepsPerDay$steps_per_day, na.rm=TRUE))
steps.median <- median(stepsPerDay$steps_per_day, na.rm=TRUE)
```
2. The mean number of steps is `r steps.mean` 
and the median number of steps is `r steps.median`.

## What is the average daily activity pattern?
1. Make a time series plot of the average number of steps over all days.  
```{r daily-activity,  echo=TRUE, eval=TRUE}
stepsByInterval <- activity %>%
        group_by(interval) %>%
        summarise(avg_steps_per_interval = mean(steps, na.rm=TRUE))

ggplot(data=stepsByInterval, aes(x=interval, y=avg_steps_per_interval)) +
        geom_line() +
        ggtitle("Average number of steps in each time interval")
max_steps <- max(stepsByInterval$avg_steps_per_interval)
max_steps_interval <- with(stepsByInterval,
                        interval[avg_steps_per_interval==max_steps])
```
2. The maximum average number of steps in any time interval, `r round(max_steps) `, 
occured in the interval starting at `r max_steps_interval`.  

## Imputing missing values
1. Calculate the missing values in the activity dataset.
```{r count_nas,  echo=TRUE, eval=TRUE}
n_missing <- sum(is.na(activity$steps))

```
There are `r n_missing ` missing values in the dataset.  
2. Impute missing values by filling in the mean value of steps for that interval.
```{r imputation,  echo=TRUE, eval=TRUE}
# Put the average number of steps per interval into the activity dataset
activity <- activity %>%
        mutate(
                avg_steps = rep(
                        stepsByInterval$avg_steps_per_interval, 
                                length(unique(activity$date))))

# if the number of steps is missing, impute the average number of steps
# rounded to the nearest whole number
activity <- activity %>%
        mutate(steps_imputed = ifelse(is.na(steps), 
                                      round(avg_steps), 
                                      steps))

```
3. Make a new dataset with the imputed values
```{r new_dataset,  echo=TRUE, eval=TRUE}
activity2 <- data.frame(steps=activity$steps_imputed, date=activity$date,
                        interval=activity$interval)
summary(activity2)
```
4. Redo the histogram and mean and median on this new dataset.
```{r redo_analysis,  echo=TRUE, eval=TRUE}
stepsPerDay2 <- activity2 %>%
        group_by(date) %>%
        summarise(steps_per_day = sum(steps, na.rm=TRUE))

ggplot(data=stepsPerDay2, aes(x=steps_per_day)) +
        geom_histogram() +
        ggtitle("Steps per day in the revised activity data")

steps.mean2 <- round(mean(stepsPerDay2$steps_per_day, na.rm=TRUE))
steps.median2 <- median(stepsPerDay2$steps_per_day, na.rm=TRUE)

```
The original mean number of steps is `r steps.mean` 
and the original median number of steps is `r steps.median`.
The revised mean number of steps is `r steps.mean2` 
and the revised median number of steps is `r steps.median2`.  

* Do these differ from the first step mean and median?  
        Yes, they do.  
* What is the impact of imputing missing data on these estimates?  
        The imputation has increase the mean more than it has increased the median.  

## Are there differences in activity patterns between weekdays and weekends?
1. Make a new factor that differentiates weekdays from weekends.
```{r split_weekends,  echo=TRUE, eval=TRUE}
weekend_days <- c("Saturday", "Sunday")
activity2$period <- ifelse(
                        weekdays(activity2$date) %in% weekend_days, "Weekend", "Weekday")

```
2. Make a panel plot that compares average steps on weekends vs. weekdays.  
```{r plot_week_day_end,  echo=TRUE, eval=TRUE}
stepsByInterval2 <- activity2 %>%
        group_by(interval, period) %>%
        summarise(avg_steps_per_interval = mean(steps, na.rm=TRUE))

ggplot(data=stepsByInterval2, aes(x=interval, y=avg_steps_per_interval, colour=period)) +
        geom_line() +
        facet_grid(period ~ .) +
        xlab("Interval") +
        ylab("Number of Steps") +
        ggtitle("Average number of steps in each time interval (revised)")




```